Staged typed type-preserving tagless ``interpreters'', \emph{finally}

### 2007-04-01 tagless/tagless-final.tex
Went over Section 2-3

\hyphenation{meta-language meta-languages}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}
\usepackage{mdwlist}

@@

We focus on embedding a typed object language into a typed metalanguage.
Types are useful as usual because they document the programmer's
intentions and can be checked early by machine.  \ccshan{Collect reasons
to type here.}  To be concrete, we use the typed object language in
Figure~\ref{fig:object} throughout this paper.  We aim not just to
interpret the object language in the narrow sense but also for
compilation, partial evaluation, and other processing.

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize*}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item even partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        including call-by-name (CBN).
\end{itemize*}
We have programmed our interpreters in OCaml (and, for staging,
\citet{metaocaml}) and standard Haskell. The complete code is available%
\footnote{The URL will appear in the final version of the paper.}
% connot give URL in the ICFP paper due to double-blind reviews
to supplement the paper. Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, our OCaml evaluator obviously
terminates except if the fixpoint combinator $\mathrm{fix}$ is present
in the object program.

@@

\section{The base language and its tagless interpreters}\label{language}

@@

\subsection{How to make encoding flexible: abstract the interpreter}
\label{encoding}
We have embedded the above language in (Meta)OCaml and Haskell.  In Haskell,
the constructors of DSL terms are functions defined as methods in a type class
Symantics. The name implies that the class interface gives the syntax for
the source language and its instances give the semantics.

\begin{code}
class Symantics repr where
  int :: Int   -> repr Int          -- int literal
  bool :: Bool -> repr Bool         -- bool literal

  lam :: (repr a -> repr b) -> repr (a->b)
  app :: repr (a->b) -> repr a -> repr b
  fix :: (repr a -> repr a) -> repr a

  add :: repr Int  -> repr Int -> repr Int
  mul :: repr Int  -> repr Int -> repr Int
  if_ :: repr Bool -> repr a   -> repr a -> repr a
  leq :: repr Int  -> repr Int -> repr Bool
\end{code}

\noindent |test1| now reads |app (lam (\x -> x)) (bool True)|,
whose inferred type is |Symantics repr => repr Bool|.
For example, we can encode the classical |power| function, as well
as the partial application |\x -> power x 7| as

\begin{code}
testpowfix () = 
  lam (\x -> fix (\self -> lam (\n ->
        if_ (leq n (int 0)) (int 1)
            (mul x (app self (add n (int (-1))))))))
testpowfix7 () = 
  lam (\x -> app (app (testpowfix ()) x) (int 7))
\end{code}
The dummy unit |()| argument above is to avoid the monomorphism
restriction and so that the type of |testpowfix| and |testpowfix7|
will remain polymorphic in |repr|. Instead of supplying this dummy
argument, we could have given the terms explicit polymorphic
signature.  We however prefer if the typechecker of the meta-language
(Haskell) infers the types of the object terms for us. We could also
avoid the |()| argument by switching off the monomophism restriction
with a compiler flag.

Note that the object language is typed. If we make a mistake
(for example, replace |int 7| with |bool True| in |testpowfix7|, the
compiler will highlight that term and complain that it couldn't match
the expected |Int| against the inferred |Bool|. A type error
also results when entering |lam (\x -> app x x)|, with the compiler
complaining about an infinite type. The type-checker of the
meta-language will flag meaningless object level terms --
before any of their evaluation. If we forget, for example, |app| in
some of the terms above, we get an error message with a precise
location. So the type-checker also flags syntactically invalid
object-level terms.\oleg{Should we perhaps move this up, in the
  intro? To justify why typed-metalanguages and the typed DSL are
  good. Or keep the explanation here but refer to it from the Introduction.}

We have also embedded our DSL in (Meta)OCaml, replacing type classes
and instances with signatures and structures. For this and the 
following sections, the following simple signature suffices:

\begin{code}
module type Symantics = sig
  type ('c,'dv) repr
  val int : int  -> ('c,int) repr
  val bool: bool -> ('c,bool) repr
  val add : ('c,int) repr-> ('c,int) repr-> ('c,int) repr
  val mul : ('c,int) repr-> ('c,int) repr-> ('c,int) repr
  val leq : ('c,int) repr-> ('c,int) repr-> ('c,bool) repr
  val if_ : ('c,bool) repr ->
             (unit -> ('c,'da) repr) ->
             (unit -> ('c,'da) repr) -> ('c,'da) repr 
  val lam : (('c,'da) repr -> ('c,'db) repr) 
          -> ('c,'da->'db) repr
  val app : ('c,'da->'db) repr
    -> ('c,'da) repr -> ('c,'db) repr
  val fix : (('c,'da->'db) repr -> ('c,'da->'db) repr) 
            -> ('c,'da->'db) repr
end;;
\end{code}

Some differences are in the additional type parameter |c| (the
environment classifier from \cite{WalidPOPL03}) to be used later when we get to
code generation (this is required by MetaOCaml as part of the type for the
code). MetaOCaml is a call-by-value language, and thus we have to 
$\eta$-expand
the fix-point combinator and encapsulate the branches of the |if_|
into thunks to prevent their evaluation.

The power example and our sample term |test1| are encoded as
\begin{code}
module EX(S: Symantics) = struct
 open S
 let test1 () = app (lam (fun x -> x)) (bool true)
 let testpowfix () = 
   lam (fun x ->fix (fun self -> lam (fun n ->
     if_ (leq n (int 0)) (fun () -> int 1)
         (fun () -> mul x (app self 
                               (add n (int (-1))))))))
 let testpowfix7 = 
    lam (fun x -> app (app (testpowfix ()) x) (int 7))
end;;
\end{code}
The dummy unit argument in |testpowfix| is an artefact of MetaOCaml
(which we will be using to compile our tests in Sections below),
again related to monorphism. To be
more precise, the environment qualifier (which appears as the |'c|
type parameter in the signature of Symantic functions) must remain
uninstantiated in order for us to `run' the code. That means the type
must be polymorhic. The value restriction then dictates that
the definitions of our object terms must look syntactically like
values. Alternatively to giving the |()| argument, we could have used
the rank-2 record types of OCaml to maintain the necessary polymorphism.

Thus the object expressions (expressions of the embedded DSL) have in
OCaml the type of a functor from |Symantics| to the desired value. This
is essentially the same as the |Symantics repr => | constraint in the case of
the embedding of the DSL in Haskell.

\subsection{Two tagless interpreters}
\label{S:interpreter-RL}
The first instantiation of the Symantics structure is an interpreter:
evaluating the object (DSL) term to its value in the meta-language. 
The following module essentially interprets (gives one-to-one
correspondence to) each operation of the object language as the
corresponding operation in the meta-language.
\begin{code}
module R  = struct
  type ('c,'dv) repr = 'dv (* absolutely no wrappers *)
  let int  (x:int)  = x
  let bool (b:bool) = b
  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then (et ()) else (ee ())

  let lam f         = f
  let app e1 e2     = e1 e2
  let fix f         = let rec self n = f self n in self
end;;
\end{code}

This interpreter is patently tagless, using neither Universal type nor
any pattern-matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. We can run our
examples by instatiating the |EX| functor with |R|.
\noindent
\begin{code}
  module EXR = EX(R);;
\end{code}

\noindent and so |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define a\\
|newtype R a = R{unR::a}| and |instance Symantics R|
with a straightforward implementation. We must stress that |R| is
a |newtype|. Although it may look like a tag, it has no run-time
representation (erased during compilation) and pattern-matching
against |R| cannot ever fail and so is operationally an identity and
compiled that way. In both cases, the absense of pattern-match
failures is obvious to the compiler, because there are no
pattern-matching. The failure to yield a value
may only result from interpreting |fix|.

\begin{prop}
If the term |e| encoded in the meta-language has type
|t|, evaluation of that term in the interpreter R will either continue
indefinitely or termninate with a value of the same type |t|.
\end{prop}

We'd like show a different interpreter: it interprets each terms of
the object language to its size (defined as the number of term
constructors). The following is slightly abbreviated code (see the
accompanying source code for the complete definition).

\begin{code}
module L = struct
  type ('c,'dv) repr = int
  let int (x:int)  = 1
  let mul e1 e2    = e1 + e2 + 1
  let if_ eb et ee = eb + et () + ee () + 1
  let lam f        = (f 0) + 1
  let app e1 e2    = e1 + e2 + 1
  let fix f        = (f 0) + 1
end;;
\end{code}

\noindent Now the expression
\begin{code}
  let module E = EX(L) in E.test1 ()
\end{code}
evaluates to |3|. This interpreter is not only tagless, it is also
total. It ``evaluates'' even seemingly divergent terms like\\
|app (fix (fun self -> self)) (int 1)|.

@@

\section{Tagless compiler (aka, staged interpreter)}\label{compiler}

We can also write a ``compiler'' from our lambda calculus directly
into OCaml code, using MetaOCaml's staging facilities. Namely,
future-stage expressions of the type |t| are represented at the
present stage as values of the type |('a,t) code| where |'a| is the
environment classifier \cite{WalidPOPL03,WalidESOP04}. Code values are created
by a \emph{bracket} form: |.<e>.| specifies the expression |e| is to be
evaluated at a future stage. The \emph{escape} |.~e|, which must occur
within a bracket, specifies that the expression |e| must be evaluated
at the current stage; its result, which must be a code value, is
spliced in into the code being built. The \emph{run} form |.! e|
evaluates the code value, which involves run-time compilation and
linking of the future-stage code. The bracket, escape and run are akin to
quasi-quotation, unquotation and |eval| of Lisp.

% this code does not have the 'sv parameter. It shows up later.
\begin{code}
module C = struct
  type ('c,'dv) repr = ('c,'dv) code
  let int (x:int)   = .<x>.
  let bool (b:bool) = .<b>.
  let add e1 e2     = .<.~e1 + .~e2>.
  let mul e1 e2     = .<.~e1 * .~e2>.
  let leq x y       = .< .~x <= .~y >.
  let if_ eb et ee = 
    .<if .~eb then .~(et () ) else .~(ee () )>.
  let lam f         = .<fun x -> .~(f .<x>.)>.
  let app e1 e2     = .<.~e1 .~e2>.
  let fix f = 
     .<let rec self n = .~(f .<self>.) n in self>.
end;;
\end{code}
This is a completely straightforward staging of
|module R|.
This compiler attempts no optimizations at all, it simply produces
residual code. For example, the evaluation of our |test1|
\begin{code}
  let module E = EX(C) in E.test1 ();;
\end{code}
gives a value with inferred type |('a, bool) C.repr| which looks
like |.<((fun x_6 -> x_6) (true))>.|. MetaOCaml can print the code
values, that is, the corresponding future-stage code.

It is readily obvious that this compiler does not suffer
any interpretive overhead whatsoever. Indeed, the
  code produced for a function is simply |fun x_6 -> x_6| and has no 
  embedded calls to the evaluator: cf. the evaluator of DSL functions
  in Section~\ref{tagproblem}, the L-line.
The resulting code obviously contains no tags and no pattern-matching.

We have also implemented this ``compiler'' in Haskell. Since Haskell
has no (convenient, typed) staging facilities, we had to emulate
them. To be precise, we defined a data type |ByteCode| with
constructors |Var|, |Lam|, |App|, |Fix|, |INT| etc. with the obvious
meaning. Unlike our staging language (which is based on higher-order
abstract syntax), our bytecode, to be realistic, uses integer-named
variables. We then define 
\begin{code}
  newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for the creation of fresh variable
names. We define the compiler by making |C| to be an instance of the
class |Symantics|. The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \textsf{ByteCode} \emph{datatype}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code given above. The
accompanying code gives the full details.

\section{Tagless partial evaluator}\label{PE}

>>>>

\hyphenation{meta-language meta-languages Meta-OCaml}
\documentclass[preprint]{sigplanconf}
\usepackage{amsmath,amssymb}
\usepackage{mathptmx}
%\usepackage{txfonts} \renewcommand{\ttdefault}{cmtt}
\usepackage[T1]{fontenc}
\usepackage{comment}
\usepackage{hyphenat}
\usepackage{url}
\usepackage{prooftree1}
\usepackage{mdwlist}

@@

We focus on embedding a typed object language into a typed metalanguage.
Types are useful as usual because they document the programmer's
intentions and can be checked early by machine (see~\S\ref{encoding}).
To be concrete, we use the typed object language in
Figure~\ref{fig:object} throughout this paper.  We aim not just to
interpret the object language in the narrow sense but also for
compilation, partial evaluation, and other processing.

@@

The term ``constructor'' functions |b|, |lam|, |app|, and so on appear
free in the encoding of an object term such as |testf1| above.  Defining
these functions differently gives rise to different interpreters, that
is, different folds on object programs.  Given the same term
representation but varying the interpreter, we can
\begin{itemize*}
    \item evaluate the term to a value in the metalanguage;
    \item measure the size or depth of the term;
    \item compile the term, with staging support such as in MetaOCaml;
    \item even partially evaluate the term, online; and
    \item transform the term to continuation\hyp passing style (CPS),
        including call-by-name (CBN).
\end{itemize*}
We have programmed our interpreters in OCaml (and, for staging,
\citet{metaocaml}) and standard Haskell. The complete code is available%
\footnote{The URL will appear in the final version of the paper.}
% connot give URL in the ICFP paper due to double-blind reviews
to supplement the paper. Our examples below switch between (Meta)OCaml
and Haskell even though we have implemented each example equivalently in
both metalanguages, because some of our claims are more obvious in one
metalanguage than the other.  For example, MetaOCaml provides
convenient, typed staging facilities.

@@

\section{The object language and its tagless interpreters}\label{language}

@@

\subsection{How to make encoding flexible: abstract the interpreter}
\label{encoding}

We embed this language in (Meta)OCaml and Haskell.  In Haskell,
the functions that construct object terms are methods in a type class
|Symantics|. The class is so named because its interface gives the syntax for
the object language and its instances give the semantics.
The parameter |repr| to the class has the kind |* -> *|.
\begin{code}
class Symantics repr where

  int  :: Int  -> repr Int
  bool :: Bool -> repr Bool

  lam :: (repr a -> repr b) -> repr (a -> b)
  app :: repr (a -> b) -> repr a -> repr b
  fix :: (repr a -> repr a) -> repr a

  add :: repr Int -> repr Int -> repr Int
  mul :: repr Int -> repr Int -> repr Int
  leq :: repr Int -> repr Int -> repr Bool
  if_ :: repr Bool -> repr a -> repr a -> repr a
\end{code}
For example, we encode the term |test1|, or $(\fun{x}x)\True$, from
\S\ref{tagproblem} above as |app (lam (\x -> x)) (bool True)|,
whose inferred type is |Symantics repr => repr Bool|.
For another example, the classical $\mathit{power}$ function is
\begin{code}
testpowfix () = 
  lam (\x -> fix (\self -> lam (\n ->
  if_ (leq n (int 0)) (int 1)
      (mul x (app self (add n (int (-1))))))))
\end{code}
and the partial application $\fun{x} \mathit{power}\;x\;7$ is
\begin{code}
testpowfix7 () = 
  lam (\x -> app (app (testpowfix ()) x)
                 (int 7))
\end{code}
The dummy argument |()| above is to avoid the monomorphism
restriction, to keep the type of |testpowfix| and |testpowfix7|
polymorphic in |repr|. Instead of supplying this dummy
argument, we could have given the terms explicit polymorphic
signatures.  We however prefer for
Haskell to infer the object types for us. We could also
avoid the dummy argument by switching off the monomophism restriction
with a compiler flag.

Because the types in the |Symantics| class reflect the object language's type
system, this representation admits only well-typed object terms for evaluation.
If we make a mistake somewhere, say replace |int 7| with |bool True| in
|testpowfix7|, Haskell will complain there that the expected type |Int| does not
match the inferred |Bool|.  Similarly, the object term $\fun{x}xx$ and its
encoding |lam (\x -> app x x)| both fail occurs-checks in type checking.
Haskell's type checker also flags syntactically invalid object terms, such as if
we forget |app| somewhere above.

\begin{figure*}
\begin{floatrule}
\begin{code}
module type Symantics = sig

  type ('c, 'dv) repr

  val int : int  -> ('c, int) repr
  val bool: bool -> ('c, bool) repr

  val lam : (('c, 'da) repr -> ('c, 'db) repr) -> ('c, 'da -> 'db) repr
  val app : ('c, 'da -> 'db) repr -> ('c, 'da) repr -> ('c, 'db) repr
  val fix : (('c, 'da -> 'db) repr -> ('c, 'da -> 'db) repr) -> ('c, 'da -> 'db) repr

  val add : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val mul : ('c, int) repr -> ('c, int) repr -> ('c, int) repr
  val leq : ('c, int) repr -> ('c, int) repr -> ('c, bool) repr
  val if_ : ('c, bool) repr -> (unit -> ('c, 'da) repr) -> (unit -> ('c, 'da) repr) -> ('c, 'da) repr

end
\end{code}
\end{floatrule}
\caption{A simple (Meta)OCaml embedding of our object language}
\label{fig:ocaml-simple}
\end{figure*}

To embed the same object language in (Meta)OCaml, we replace the type
class |Symantics| and its instances by a module signature |Symantics|
and its implementations.  Figure~\ref{fig:ocaml-simple} shows a simple
signature that suffices for this and the following sections.
Two differences between this signature and the |Symantics| class in
Haskell are: the additional type parameter |c|, an
environment classifier \citep{WalidPOPL03} required by MetaOCaml for
code generation in~\S\ref{compiler}; and the $\eta$-expanded type for
|fix| and thunk types in |if_| due to OCaml's being a call-by-value
language.

The functor below encodes our examples |test1| and $\mathit{power}$.
\begin{code}
module EX(S: Symantics) = struct
  open S
  let test1 () =
    app (lam (fun x -> x)) (bool true)
  let testpowfix () = 
    lam (fun x -> fix (fun self -> lam (fun n ->
    if_ (leq n (int 0)) (fun () -> int 1)
        (fun () -> mul x
          (app self (add n (int (-1))))))))
  let testpowfix7 = 
     lam (fun x -> app (app (testpowfix ()) x)
                       (int 7))
end;;
\end{code}
The dummy argument to |test1| and |testpowfix| is an artefact of
MetaOCaml, again related to monomorphism: in order for us to run a
piece of generated code, it must be polymorphic in its environment
classifier (the type variable |'c| in Figure~\ref{fig:ocaml-simple}).
The value restriction then dictates that
the definitions of our object terms must look syntactically like
values. Alternatively to giving the |()| argument, we could have used
the rank-2 record types of OCaml to maintain the necessary polymorphism.

Thus we represent an object expression in
OCaml as a functor from |Symantics|. This
is essentially the same as the constraint |Symantics repr =>| in the
Haskell embedding.

\subsection{Two tagless interpreters}
\label{S:interpreter-RL}

Having abstracted our term representation over the interpreter, we are
now ready to present a series of interpreters.  Each interpreter is an
instance of the |Symantics| class in Haskell and a module implementing
the |Symantics| signature in MetaOCaml.

The first interpreter evaluates the object term to its value in the
metalanguage.  The module below interprets each object\hyp language
operation as the corresponding metalanguage operation.
\begin{code}
module R = struct

  (* absolutely no wrappers *)
  type ('c,'dv) repr = 'dv

  let int  (x:int)  = x
  let bool (b:bool) = b

  let lam  f        = f
  let app  e1 e2    = e1 e2
  let fix  f        =
    let rec self n = f self n in self

  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee ()

end
\end{code}

As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding} with this |R| module.
\begin{code}
module EXR = EX(R)
\end{code}
Thus |EXR.test1 ()| evaluates to the untagged boolean value |true|.
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is compiled to no-op.
In both OCaml and Haskell, it is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
That is, the proposition below follows immediately from the soundness of
the metalanguage's type system.
\begin{prop}
If the term |e| encoded in the metalanguage has type~|t|,
then evaluating~|e| in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~|t|.
\end{prop}

For variety, we show another interpreter, which measures the \emph{size}
of each object term, defined as the number of term
constructors. The following is slightly abbreviated code (see the
accompanying source code for the complete definition).
\begin{code}
module L = struct
  type ('c,'dv) repr = int
  let int (x:int)  = 1
  let lam f        = f 0 + 1
  let app e1 e2    = e1 + e2 + 1
  let fix f        = f 0 + 1
  let mul e1 e2    = e1 + e2 + 1
  let if_ eb et ee = eb + et () + ee () + 1
end;;
\end{code}
Now the expression
\begin{code}
let module E = EX(L) in E.test1 ()
\end{code}
evaluates to |3|. This interpreter is not only tagless but also
total. It ``evaluates'' even seemingly divergent terms like
\begin{code}
app (fix (fun self -> self)) (int 1)
\end{code}

@@

\section{A tagless compiler (or, a staged interpreter)}\label{compiler}

We can also compile our object language
into OCaml code using MetaOCaml's staging facilities. MetaOCaml
represents future-stage expressions of type |t| at the
present stage as values of type |('c,t) code| where |'c| is the
environment classifier \cite{WalidPOPL03,calcagno-ml-like}. Code values are created
by a \emph{bracket} form |.<e>.|, which specifies that the expression |e| is to be
evaluated at a future stage. The \emph{escape} |.~e| must occur
within a bracket and specifies that the expression |e| must be evaluated
at the current stage; its result, which must be a code value, is
spliced into the code being built by the enclosing bracket. The \emph{run} form |.! e| evaluates
the future-stage code value~|e| by compiling and linking it at run-time.
The bracket, escape, and run are akin to
quasi-quotation, unquotation, and |eval| of Lisp.

Inserting brackets and escapes appropriately into the evaluating
interpreter~|R| above yields the compiling interpreter~|C| below.
% this code does not have the 'sv parameter. It shows up later.
\begin{code}
module C = struct

  type ('c,'dv) repr = ('c,'dv) code

  let int (x:int)   = .<x>.
  let bool (b:bool) = .<b>.

  let lam f         = .<fun x -> .~(f .<x>.)>.
  let app e1 e2     = .<.~e1 .~e2>.
  let fix f = 
    .<let rec self n = .~(f .<self>.) n in self>.

  let add e1 e2     = .<.~e1 + .~e2>.
  let mul e1 e2     = .<.~e1 * .~e2>.
  let leq x y       = .<.~x <= .~y>.
  let if_ eb et ee = 
    .<if .~eb then .~(et ()) else .~(ee ())>.

end
\end{code}
This is a completely straightforward staging of
|module R|.
This compiler simply produces
residual code with no optimization. For example, evaluating our |test1|
\begin{code}
let module E = EX(C) in E.test1 ()
\end{code}
gives the code value |.<(fun x_6 -> x_6) (true)>.|
of inferred type |('c, bool) C.repr|.

This compiler does not incur
any interpretive overhead: the
code produced for $\fun{x}x$ is simply |fun x_6 -> x_6| and contains no 
  call to the evaluator, unlike the recursive calls to |eval0| and
  |eval| in the |L e| lines in \S\ref{tagproblem}.
The resulting code obviously contains no tags and no pattern matching.

We have also implemented this compiler in Haskell. Since Haskell
has no (convenient, typed) staging facilities, we had to emulate
them. To be precise, we defined a data type |ByteCode| with
constructors such as |Var|, |Lam|, |App|, |Fix|, and |INT|.
Whereas our representation of object terms uses higher-order
abstract syntax, our bytecode uses integer-named
variables to be realistic. We then define 
\begin{code}
newtype C t = C (Int -> (ByteCode t, Int)) 
\end{code}
where |Int| is the counter for creating fresh variable
names. We define the compiler by making |C| an instance of the
class |Symantics|. The implementation\footnote{The implementation uses
GADTs because we also wanted to write a typed interpreter for 
the \textsf{ByteCode} \emph{datatype}.} is quite similar (but slightly more
verbose) than the corresponding MetaOCaml code given above. The
accompanying code gives the full details.

\section{A tagless partial evaluator}\label{PE}
